---
title: "Dynamics with two variables"
author: "Sam Muir & Melissa Widas"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(sensitivity)
library(tidyverse)
library(deSolve)
library(lhs)
library(purrr)
library(ggpubr)
library(patchwork)
source("lotvmodK.R")
lotvmodK
```

```{r}
# Initial conditions
currpop <- c(prey = 10, pred = 1)

# Set parameter list
pars <- c(rprey = 0.5, K = 100, alpha = 0.3, beta = 0.05, eff = 0.2, pmort = 0.2)

# Times when you want to evaluate
days <- seq(from = 1, to = 500, by = 1)

# run our differential equation solver
res = ode(func=lotvmodK, y=currpop, times=days, parms=pars, thresh = 5)
```

```{r}
# rearrange for plotting
resl = as.data.frame(res) %>% 
  pivot_longer(-time, names_to="species", values_to="pop")

# graph both populations over time
ggplot(resl, aes(time, pop, col=species)) +
  geom_line() +
  theme_linedraw() +
  labs(x = "Time", y = "Population")
```


Stability
```{r}
pars_stab <- data.frame(rprey = 0.95, K = 2000, alpha = 0.01, beta = 0.05, eff = 0.4, pmort = 0.4)

# now lets try initial conditions that will be stable
preyi = with(pars_stab, pmort/(eff*alpha))
predi = with(pars_stab, rprey / (alpha + beta) * (1-preyi/K))

preyi
predi

# times when you want to evaluate
days = seq(from=1,to=500)

# lets first see what happens when we start with 1 of each
currpop=c(prey=1, pred=1)
# run our differential equation solver
res = ode(func=lotvmodK, y=currpop, times=days, parms=pars_stab, thresh = 20)
# extract the results
res_smallstart = as.data.frame(res) %>% 
  gather(key="animal", value="pop",-time)
# graph both populations over time
ggplot(res_smallstart, aes(time, pop, col=animal)) +
  geom_line() +
  theme_linedraw() +
  labs(x = "Time", y = "Population")


# lets first see what happens when we start our estimates of stable populations
stablepop = c(prey=preyi, pred=predi)
res = ode(func=lotvmodK, y=stablepop, times=days, parms=pars_stab, thresh = 20)

# extract the results
res_stablestart = as.data.frame(res) %>% 
  gather(key="animal", value="pop",-time)

# graph both populations over time
ggplot(res_stablestart, aes(time, pop, col=animal)) + 
  geom_line() +
  theme_linedraw() +
  labs(x = "Time", y = "Population")

```

We're defining stability as when the predator and prey populations do not change over time after 50 years. 

Exploring Different Hunting Levels

```{r}
pars_hunt1 <- data.frame(rprey = 0.95, K = 2000, alpha = 0.01, beta = 0.1, eff = 0.4, pmort = 0.4)

# now lets try initial conditions that will be stable
preyi = with(pars_hunt1, pmort/(eff*alpha))
predi = with(pars_hunt1, rprey / (alpha + beta) * (1-preyi/K))

# times when you want to evaluate
days = seq(from=1,to=200)

# lets first see what happens when we start with 1 of each
currpop=c(prey=preyi, pred=predi)
# run our differential equation solver
res = ode(func=lotvmodK, y=currpop, times=days, parms=pars_hunt1, thresh = 20)
# extract the results
res_hunt1 = as.data.frame(res) %>% 
  gather(key="animal", value="pop",-time)
# graph both populations over time
hunt1 <- ggplot(res_hunt1, aes(time, pop, col=animal)) +
  geom_line() +
  theme_linedraw() +
  labs(x = "Time", y = "Population", title = "Beta 0.1 with a threshold of 20")
```

```{r}
pars_hunt2 <- data.frame(rprey = 0.95, K = 2000, alpha = 0.01, beta = 0.05, eff = 0.4, pmort = 0.4)

# now lets try initial conditions that will be stable
preyi = with(pars_hunt2, pmort/(eff*alpha))
predi = with(pars_hunt2, rprey / (alpha + beta) * (1-preyi/K))

# times when you want to evaluate
days = seq(from=1,to=200)

# lets first see what happens when we start with 1 of each
currpop=c(prey=preyi, pred=predi)
# run our differential equation solver
res = ode(func=lotvmodK, y=currpop, times=days, parms=pars_hunt2, thresh = 5)
# extract the results
res_hunt2 = as.data.frame(res) %>% 
  gather(key="animal", value="pop",-time)
# graph both populations over time
hunt2 <- ggplot(res_hunt2, aes(time, pop, col=animal)) +
  geom_line() +
  theme_linedraw() +
  labs(x = "Time", y = "Population", title = "Beta 0.05 with a threshold of 5")
```

```{r}
pars_hunt3 <- data.frame(rprey = 0.95, K = 2000, alpha = 0.01, beta = 0.05, eff = 0.4, pmort = 0.4)

# now lets try initial conditions that will be stable
preyi = with(pars_hunt3, pmort/(eff*alpha))
predi = with(pars_hunt3, rprey / (alpha + beta) * (1-preyi/K))

# times when you want to evaluate
days = seq(from=1,to=200)

# lets first see what happens when we start with 1 of each
currpop=c(prey=preyi, pred=predi)
# run our differential equation solver
res = ode(func=lotvmodK, y=currpop, times=days, parms=pars_hunt3, thresh = 100, method = "ode45")
# extract the results
res_hunt3 = as.data.frame(res) %>% 
  gather(key="animal", value="pop",-time)
# graph both populations over time
hunt3 <- ggplot(res_hunt3, aes(time, pop, col=animal)) +
  geom_line() +
  theme_linedraw() +
  labs(x = "Time", y = "Population", title = "Beta 0.05 with a threshold of 100")

hunt3
```

```{r}
pars_hunt4 <- data.frame(rprey = 0.95, K = 2000, alpha = 0.01, beta = 0.05, eff = 0.4, pmort = 0.4)

# now lets try initial conditions that will be stable
preyi = with(pars_hunt4, pmort/(eff*alpha))
predi = with(pars_hunt4, rprey / (alpha + beta) * (1-preyi/K))

# times when you want to evaluate
days = seq(from=1,to=200)

# lets first see what happens when we start with 1 of each
currpop=c(prey=preyi, pred=predi)
# run our differential equation solver
res = ode(func=lotvmodK, y=currpop, times=days, parms=pars_hunt4, thresh = 50)
# extract the results
res_hunt4 = as.data.frame(res) %>% 
  gather(key="animal", value="pop",-time)
# graph both populations over time
hunt4 <- ggplot(res_hunt4, aes(time, pop, col=animal)) +
  geom_line() +
  theme_linedraw() +
  labs(x = "Time", y = "Population", title = "Beta 0.05 with a threshold of 50")

hunt4
```

```{r}
pars_hunt5 <- data.frame(rprey = 0.95, K = 2000, alpha = 0.01, beta = 0.34, eff = 0.4, pmort = 0.4)

# now lets try initial conditions that will be stable
preyi = with(pars_hunt5, pmort/(eff*alpha))
predi = with(pars_hunt5, rprey / (alpha + beta) * (1-preyi/K))

# times when you want to evaluate
days = seq(from=1,to=200)

# lets first see what happens when we start with 1 of each
currpop=c(prey=preyi, pred=predi)
# run our differential equation solver
res = ode(func=lotvmodK, y=currpop, times=days, parms=pars_hunt5, thresh = 20)
# extract the results
res_hunt5 = as.data.frame(res) %>% 
  gather(key="animal", value="pop",-time)
# graph both populations over time
hunt5 <- ggplot(res_hunt5, aes(time, pop, col=animal)) +
  geom_line() +
  theme_linedraw() +
  labs(x = "Time", y = "Population", title = "Beta 0.34 with a threshold of 20")

hunt5
```

```{r}
pars_hunt6 <- data.frame(rprey = 0.95, K = 2000, alpha = 0.01, beta = 0.34, eff = 0.4, pmort = 0.4)

# now lets try initial conditions that will be stable
preyi = with(pars_hunt6, pmort/(eff*alpha))
predi = with(pars_hunt6, rprey / (alpha + beta) * (1-preyi/K))

# times when you want to evaluate
days = seq(from=1,to=200)

# lets first see what happens when we start with 1 of each
currpop=c(prey=preyi, pred=predi)
# run our differential equation solver
res = ode(func=lotvmodK, y=currpop, times=days, parms=pars_hunt6, thresh = 50)
# extract the results
res_hunt6 = as.data.frame(res) %>% 
  gather(key="animal", value="pop",-time)
# graph both populations over time
hunt6 <- ggplot(res_hunt6, aes(time, pop, col=animal)) +
  geom_line() +
  theme_linedraw() +
  labs(x = "Time", y = "Population", title = "Beta 0.34 with a threshold of 50")

hunt6
```

```{r}
pars_hunt7 <- data.frame(rprey = 0.95, K = 2000, alpha = 0.01, beta = 0.3, eff = 0.4, pmort = 0.4)

# now lets try initial conditions that will be stable
preyi = with(pars_hunt7, pmort/(eff*alpha))
predi = with(pars_hunt7, rprey / (alpha + beta) * (1-preyi/K))

# times when you want to evaluate
days = seq(from=1,to=200)

# lets first see what happens when we start with 1 of each
currpop=c(prey=preyi, pred=predi)
# run our differential equation solver
res = ode(func=lotvmodK, y=currpop, times=days, parms=pars_hunt7, thresh = 20)
# extract the results
res_hunt7 = as.data.frame(res) %>% 
  gather(key="animal", value="pop",-time)
# graph both populations over time
hunt7 <- ggplot(res_hunt7, aes(time, pop, col=animal)) +
  geom_line() +
  theme_linedraw() +
  labs(x = "Time", y = "Population", title = "Beta 0.3 with a threshold of 20")
```

```{r}
pars_hunt8 <- data.frame(rprey = 0.95, K = 2000, alpha = 0.01, beta = 0.5, eff = 0.4, pmort = 0.4)

# now lets try initial conditions that will be stable
preyi = with(pars_hunt8, pmort/(eff*alpha))
predi = with(pars_hunt8, rprey / (alpha + beta) * (1-preyi/K))

# times when you want to evaluate
days = seq(from=1,to=200)

# lets first see what happens when we start with 1 of each
currpop=c(prey=preyi, pred=predi)
# run our differential equation solver
res = ode(func=lotvmodK, y=currpop, times=days, parms=pars_hunt8, thresh = 20, method = "ode45")
# extract the results
res_hunt8 = as.data.frame(res) %>% 
  gather(key="animal", value="pop",-time)
# graph both populations over time
hunt8 <- ggplot(res_hunt8, aes(time, pop, col=animal)) +
  geom_line() +
  theme_linedraw() +
  labs(x = "Time", y = "Population", title = "Beta 0.5 with a threshold of 20")

hunt8
```


Beta set at 0.34 and vary threshold
```{r}
hunt5+hunt6
```

Beta set at 0.05 and vary threshold
```{r}
hunt3 + hunt4 / hunt2
```

Beta varied and threshold set at 20
```{r}
hunt1/hunt7/hunt8
```

**We tested hunting thresholds between 5 and 100, as well as hunting rates from 0.05 to 0.34. All of these combinations resulted in stable predator and prey populations after approximately 150 years. The combination of a hunting rate of 0.05 and a threshold level of 100 resulted in the bade ODE solver failing to run for the entirety of the time sequence. However, when the ODE solver was changed to ODE45 the combination ran successfully for the entirety of the time sequence. This combination of hunting rate and minimum prey threshold stabilized the fastest out of all of the combinations. Overall as our hunting rate increases the prey and predator populations stabilize more quickly. Our recommendation based off of this exploration with a stability defined as the predator and prey populations not changing after 50 years would be a hunting rate of 0.05 and a minimum prey threshold of 100.**
