---
title: "Dynamic Modeling: Sobol ODE"
authors: "Sam Muir and Melissa Widas"
date: "9 May 2024"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(deSolve)
library(sensitivity)
```

# 1. Implement model in R
```{r}
# read in the function
source("dforestgrowth.R")
```

# 2. Run the model for 300 years (using the ODE solver) starting with an initial forest size of 10 kg/C, and using the following parameters:       
-   canopy closure threshold of 50 kgC
-   K = 250 kg C (carrying capacity)
-   r= 0.01 (exponential growth rate before before canopy closure)
-   g = 2 kg/year (linear growth rate after canopy closure)

```{r}
# define time, initial forest size, and threshold
time = seq(1, 300)
Cinitial = 10
thresh = 50

# get params into a list
parms = list(K = 250, r = 0.01, g = 2)
result = ode(y = Cinitial, 
             times = time,
             func = dforestgrowth, # forest growth function
             parms = parms,
             thresh = thresh)
# get results as a dataframe
result <- as.data.frame(result)
```

# 3. Graph the Results
```{r}
ggplot(result, aes(time, `1`)) +
  geom_point() +
  theme_linedraw() +
  labs(x = "Time (years)", y = "Forest Size (kg/C)")
```

# 4. Run sobol global
Run a sobol global sensitivity analysis (vary all parameters at the same time) that explores how the estimated maximum forest size (e.g maximum of C 300 years), varies with these parameters:

-   pre canopy closure growth rate (r)      
-   post-canopy closure growth rate (g)       
-   canopy closure threshold and carrying capacity (K)      

```{r}
# generate the sobol parameters; standard deviation of 10% of mean value
np = 2000
K = rnorm(mean=250, sd=250*0.01, n=np)
r = rnorm(mean=0.01, sd=0.01*0.01, n=np)
g = rnorm(mean = 2, sd = 2*0.01, n = np)
X1 = cbind.data.frame(r=r, K=K, g = g)

# repeat for second set of samples
K = rnorm(mean=250, sd=250*0.01, n=np)
r = rnorm(mean=0.01, sd=0.01*0.01, n=np)
g = rnorm(mean = 2, sd = 2*0.01, n = np)
X2 = cbind.data.frame(r=r, K=K, g = g)

# fix any negative values
X1 = X1 %>% map_df(pmax, 0.0)
X2 = X2 %>% map_df(pmax, 0.0)

# run sobol model
sens_P = sobolSalt(model = NULL,X1, X2, nboot = 300)

# add column names 
colnames(sens_P$X) = c("r","K", "g")
```

```{r}
# gets results for 300 years (evaluating every year)
thresh = 50
simtimes = seq(from=1, to=300)
parms = list(r=sens_P$X[1,"r"], 
             K=sens_P$X[1,"K"], 
             g=sens_P$X[1,"g"])

# run ode solver
result = ode(y = Cinitial, 
             times = simtimes,
             func = dforestgrowth, # growth function
             parms = parms,
             thresh = thresh)

result = as.data.frame(result)

# rename the column names
colnames(result)=c("time","C")
```

Compute metrics and wrapper functions

```{r}
# turn computing our metrics into a function
compute_metrics = function(result, thresh) {
  
  maxgrowth = max(result$C)
  
  idx = which(result$C > thresh)[1] # select first year above threshold
  idx = ifelse(is.na(idx), # if no results from greater than threshold
               length(result$C), # set index to the length of results 
               idx) # other wise keep
  
  threshyear = result$time[idx]
  
return(list(maxgrowth=maxgrowth, threshyear=threshyear))
}
```

```{r}
# try it on our first parameter set
p_wrapper = function(r, K, g, Cinitial, simtimes, odefunc, metricfunc, thresh) {
    parms = list(r=r, K=K, g=g)
    result = ode(y=Cinitial, 
                 times=simtimes, 
                 func=odefunc, 
                 parms=parms,
                 thresh = thresh) 
    colnames(result)=c("time","C")
  # get metrics
  metrics=metricfunc(as.data.frame(result), thresh=100)
  return(metrics)
}
```

Apply wrapper, growth function, and compute metrics
```{r}
all = as.data.frame(sens_P$X) %>% 
  pmap(p_wrapper, 
       thresh=50,
       Cinitial=Cinitial, 
       simtimes=simtimes,
       odefunc=dforestgrowth,
       metricfunc=compute_metrics)

# extract results from all into a data frame
allres = all %>% 
  map_dfr(`[`,c("maxgrowth","threshyear"))
```

# 5. Graph results of sensitivity analysis as a box plot of maximum forest size and record the two Sobol indices (S and T).
```{r}
# Boxplots
tmp = allres %>% 
  pivot_longer(cols=everything(),
               names_to="metric", 
               values_to="value")

ggplot(tmp, aes(metric, value, color = metric)) + 
  geom_boxplot() + 
  theme_linedraw()
```

```{r}
# Get sobol indices ----
sens_P_maxgrowth = sensitivity::tell(sens_P,
                                  allres$maxgrowth)

# first-order indices (main effect without co-variance)
rownames(sens_P_maxgrowth$S)=c("r","K", "g")
sens_P_maxgrowth$S

# total sensitivity index -note that this partitions the output variance 
rownames(sens_P_maxgrowth$T)=c("r","K", "g")
sens_P_maxgrowth$T


# max year ----
sens_P_threshyear = sensitivity::tell(sens_P,allres$threshyear)
# first-order indices (main effect without co-variance)
rownames(sens_P_threshyear$S)=c("r","K", "g")
sens_P_threshyear$S

# total sensitivity index -note that this partitions the output variance - so values sum to 1
rownames(sens_P_threshyear$T)=c("r","K", "g")
sens_P_threshyear$T
```

The main effect for the parameters are 0.32 (r), 0.36 (K), and 0.23 (g). The sensitivity for max growth rate for these parameters are 0.36 (r), 0.39 (K), and 0.24 (g).

The main effect on threshold year for the parameters are 0.92 (r), -0.04 (K), and 0.003 (g. The sensitivity for max growth rate for these parameters are 0.98 (r), 0.03 (K), and 0.07 (g).

# 6. Discuss meaning of results of simulation (For example think about how what parameters climate change might influence).
